Estimating the Initial Susceptible Fraction in Dengue Dynamics

### 2015-01-30 draft_lm1.tex
further edits

The attack ratio of an epidemic is define by the number of infections divided 
by the size of the population at risk.

@@

Figure \ref{fig:rtseries}  shows the $R_t$ series, according to 
(\ref{eq:Rtestimate}). 
It can be seen that the inter-epidemic periods are characterized by $R_t$ being 
indistinguishable from 1. Due to the instrinsic variability of the \rr series, 
the exmination of its credible intervals is essential to identify periods 
of sustained transmission. The wider credible intervals for the interepidemic 
periods are due to the scarsity of cases. Our methodology to derive the 
probability distribution for \rr is an original contribution applicable to the 
analysis of any incidence time-series of  any transmissible disease.

@@

Figure \ref{Fig:S0}, shows the model from (\ref{eq:model}) fitted to the data. 
In it we can see that the susceptibles series in each epidemic starts at the 
estimated level of $S_0$.
The proportion of susceptibles may seem low, but we 
must remember that these estimates are being affected by an unknown 
underreporting factor, which experts suggest is somewhere between 5 and 10, 
i.e. for every case observed there are 5 or 10 unobserved.
Since this underreporting affects both the numerator and denominator of 
(\ref{eq:AR}), its effects should cancel out, giving us an unbiased attack ratio 
estimate.
One other possible source of bias which would lead to the 
underestimation of $S_0$ could come from a significant part of the population 
not being exposed to the disease.
However, as we can see in figure \ref{fig:mapas} , the disease spreads pretty 
much homogeneously over 
the entire city (in the last four epidemics, at least).

@@

Table \ref{tab:AR} contains the attack ratios and medians of the $S_0$ 
estimated for each epidemic/outbreak. It is interesting to notice that the 
larger epidemics, in terms of total number of cases are not the one with the 
greater attack ratios. This stresses the importance of knowing the 
immunological structure of the population in order to truly assess the impact 
of any epidemic.

@@

An advantage of the Bayesian approach is that one can devise prior 
distributions for $\theta_t$ taking advantage of the intuitive parameterization 
and flexibility of the beta family of distributions.
Prior elicitation can also be done for \rr and the hyperparameters directly 
plugged into the prior for $\theta_t$. 
One can, for example, choose prior mean and variance for \rr and find $a_0$ 
and $b_0$ that satisfy those conditions.
Let $m_0$ and $v_0$ be the prior expectation and variance for $R_t$. 
After some tedious algebra one finds
\begin{align}
\label{seq:elicitation}
a_0 &= \frac{m_0v_0 + m_0^3 + m_0^2}{v_0} \\
b_0 &= \frac{2v_0 + m_0^2 + m_0}{v_0}
\end{align}
If one wants only to specify $m_0$ and a coefficient of variation $c$ 
\footnote{$c = \sqrt{v_0}/ m_0$.} for $R_t$ \textit{a priori}, some less boring 
algebra gives:
\begin{align}
\label{seq:elicitationcv}
a_0 &= \frac{m_0^3c^2 + m_0^3 + m_0^2}{m_0^2c^2} \\
b_0 &= \frac{2m_0^2c^2 + m^2 + m}{m_0^2c^2}
\end{align}

@@

\bibliography{lm1}
\section*{Figure Legends}
% This section is for figure legends only, do not include
% graphics in your manuscript file.
%
%\begin{figure}
%\caption{
%{\bf Bold the first sentence.}  Rest of figure caption.  
%}
%\label{Figure_label}
%\end{figure}
\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{plots/concat_SI.png}
\end{center}
\caption{
{\bf Susceptibles and Infectious posterior curves.}  The curves were estimated 
only for the periods where $R_t> 1$.  The susceptible curves in the top panel 
reflect the prevalence of fraction of susceptibles to circulating strain(s) for 
each epidemic/outbreak. In the lower panel see the posterior distribution 
of infectious curves, represented by its Median and 95\% credible interval. The 
credible intervals are very narrow, and can be hard to distiguish from the 
median line. 
Scatter plot shows the observed cases. Prevalences are scaled as fractions 
of the entire population.
}
\label{Fig:S0}
\end{figure}

>>>>

The attack ratio of an epidemic is defined by the number of infections divided 
by the size of the population at risk.

@@

Figure \ref{fig:rtseries}  shows the $R_t$ series, according to 
(\ref{eq:Rtestimate}). 
It can be seen that the inter-epidemic periods are characterized by $R_t$ being 
indistinguishable from 1. Due to the intrinsic variability of the \rr series, 
the examination of its credible intervals is essential to identify periods 
of sustained transmission. The wider intervals between epidemics are due to the 
scarcity of cases during these periods. Our methodology to derive the 
probability distribution of \rr is an original contribution applicable to the 
analysis of any incidence time-series of  any transmissible disease.

@@

Figure \ref{Fig:S0}, shows the model from (\ref{eq:model}) fitted to the data. 
In it we can see that the susceptibles series in each epidemic starts at the 
estimated level of $S_0$.
The proportion of susceptibles may seem low, but we 
must remember that these estimates are being affected by an unknown 
underreporting factor, which experts suggest is somewhere between 5 and 10, 
i.e. for every case observed there are 5 or 10 unobserved.
Since this underreporting affects both the numerator and denominator of 
(\ref{eq:AR2}), its effects should cancel out, giving us an unbiased attack 
ratio 
estimate.
One other possible source of bias which would lead to the 
underestimation of $S_0$ could come from a significant part of the population 
not being exposed to the disease.
However, as we can see in figure \ref{fig:mapas} , the disease spreads pretty 
much homogeneously over 
the entire city (in the last four epidemics, at least).

@@

Table \ref{tab:AR} contains the attack ratios and medians of the $S_0$ 
estimated for each epidemic/outbreak. It is interesting to notice that the 
larger epidemics, in terms of peak size are not the one with the 
greater attack ratios. This stresses the importance of knowing the 
immunological structure of the population in order to truly assess the impact 
of any epidemic.

@@

An advantage of the Bayesian approach is that one can devise prior 
distributions for $\theta_t$ taking advantage of the intuitive parameterization 
and flexibility of the beta family of distributions.
Prior elicitation can also be done for \rr and the hyper-parameters directly 
plugged into the prior for $\theta_t$. 
One can, for example, choose prior mean and variance for \rr and find $a_0$ 
and $b_0$ that satisfy those conditions.
Let $m_0$ and $v_0$ be the prior expectation and variance for $R_t$. 
After some tedious algebra one finds
\begin{align}
\label{seq:elicitation}
a_0 &= \frac{m_0v_0 + m_0^3 + m_0^2}{v_0} \\
b_0 &= \frac{2v_0 + m_0^2 + m_0}{v_0}
\end{align}
If one wants only to specify $m_0$ and a coefficient of variation $c$ 
\footnote{$c = \sqrt{v_0}/ m_0$.} for $R_t$ \textit{a priori}, some less boring 
algebra gives:
\begin{align}
\label{seq:elicitationcv}
a_0 &= \frac{m_0^3c^2 + m_0^3 + m_0^2}{m_0^2c^2} \\
b_0 &= \frac{2m_0^2c^2 + m^2 + m}{m_0^2c^2}
\end{align}

@@

\bibliography{lm1}
\section*{Figure Legends}
% This section is for figure legends only, do not include
% graphics in your manuscript file.
%
%\begin{figure}
%\caption{
%{\bf Bold the first sentence.}  Rest of figure caption.  
%}
%\label{Figure_label}
%\end{figure}
\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{plots/concat_SI.png}
\end{center}
\caption{
{\bf Susceptibles and Infectious posterior curves.}  The curves were estimated 
only for the periods where $R_t> 1$.  The susceptible curves in the top panel 
reflect the prevalence of fraction of susceptibles to circulating strain(s) for 
each epidemic/outbreak. In the lower panel see the posterior distribution 
of infectious curves, represented by its Median and 95\% credible interval. The 
credible intervals are very narrow, and can be hard to distinguish from the 
median line. 
Scatter plot shows the observed cases. Prevalences are scaled as fractions 
of the entire population.
}
\label{Fig:S0}
\end{figure}
